mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2023-01-14 rj/mathml.Rmd
finished the vignette

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-12-09"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the RMarkdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  few working examples illustrate the use of the package in a scientific report
  and in on an interactive web page.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble:
- \usepackage{longtable}
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

The paper is organized as follows. Section\ 2 presents the technical background
of the package, including the two main classes of Prolog rules for translating
R\ objects to mathematical expressions. Section\ 3 illustrates the main features
of the \pkg{mathml} package, potential issues and workarounds using examples
from the day-to-day perspective of a user. Sections\ 4 and\ 5 present two case
studies for the use of the package in a scientific report and 
a **ShinyApp web page**. Section\ 6 concludes with a discussion and ideas for
further development. A complete list of the R\ elements covered by the package
is found in the Appendix.

*shinyapp missing so far

@@

hook(successes, k)
hook(Ntotal, N)
hook(prob, pi)
mathout(term)
```

To improve the situation, \pkg{mathml} provides a simple hook that can be used
to replace elements (e.g., verbose variable names) of the code by concise
mathematical symbols, as illustrated in the example. Care should be taken to
avoid recursive hooks such as `hook(s, s["A"])` that endlessly replace
the \(s\) from \(s_{\mathrm{A}}\) as
in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).

more stuff: math/2, ml/3, jax/3

@@

mathout(call("=", quote(E(T[tau])), ET))
```

@@

```{r, results="asis", echo=FALSE}
lsq = function(tau, m, s, n)
{
  invisible(theta <- c(mu_A=0.53, sigma_A=4.3, mu_B=1.34, sigma_B=11.7, mu_M=160))
  optim(theta, gof, tau=tau, m=m, s=s, n=n, hessian=TRUE)
}

@@

```{r params, echo=FALSE}
fit <- lsq(tau, m, s, n)
theta <- fit$par
se = sqrt(diag(solve(fit$hessian)))
cilo = qnorm(0.025, theta, se)
ciup = qnorm(0.975, theta, se)

@@

This package expands the possibilities of R\ by allowing calculations and graphical rendition at the same time. This makes it easier for the user to notice writing errors and correct them smoothly. Building on current features of R\ and existing packages for displaying formulas in R\ (e.g. Murrell and Ihaka [-@murrell2000], Allaire et al. [-@allaire2018]), the \pkg{mathml} package bridges the gap between computational needs, presentation of results, and their reproducibility. To do so, it uses the Prolog's grammar rules and translates the R\ expressions to MathML. Then, two possible solutions for typesetting are available to export R\ Markdown documents in different formats or ShinyApp webpages: MathML and MathJax. A further, not-yet-mentioned, use of \pkg{mathml} is related to presentations  made using R Markdown. By adopting \pkg{mathml} researchers or teachers can use the \R markdown to both conduct analysis and show results, smoothing the process and uniforming the graphical output.

The article and Appendix show examples of how the package can be used and what operations are currently possible. Additionally, \pkg{mathml} allows for user's custom hooks to provide a better definition of the needed functions.

In its current version, however, \pkg{mathml} has some limitations. For example, it is still not possible to use the functions of \pkg{mathml} for writing intratextual formulas, while it is still more convenient to adopt the usual LateX notation. However, we do not consider this to be a serious limitation, since the most productive use of \pkg{mathml} still comes from more complicated calculations. Nevertheless, in the future versions of this package we will address this issue, to broaden its spectrum of possible applications.

Also: no linebreaks in long equations. Mostly because it is not properly
supported by MathML renderers, and also in Latex, package "breqn" (insert
reference) is mostly a proof of concept. No support for `\begin{align*}` with
manual linebreaks etc.

\pkg{mathml} is available for R\ version (…) and later, and can be easily installed using the usual `install.packages("mathml")`. The source code of the package is found at (…) including installation instructions for Unix, Windows and macOS.

## Mistakes

@@

# Acknowledgments

## Appendix

@@

>>>>

---
title: "mathml: Translate R expressions to MathML and LaTeX/MathJax"
date: "2022-12-09"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the RMarkdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package allows to use the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks allow to extend the package by
  mapping specific variables or functions to new MathML and LaTeX entities. A
  few working examples illustrate the use of the package in a scientific report.
draft: true
author:  
  # see ?rjournal_article for more information
  - name: Matthias Gondan
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: University of Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: RJreferences.bib
preamble:
- \usepackage{longtable}
- \usepackage{cancel}
- \usepackage{amsmath}
Vignette: >
  %\VignetteIndexEntry{mathml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

@@

The vignette is organized as follows. Section\ 2 presents the technical
background of the package, including the two main classes of Prolog rules for
translating R\ objects to mathematical expressions. Section\ 3 illustrates the
main features of the \pkg{mathml} package, potential issues and workarounds
using examples from the day-to-day perspective of a user. Section\ 4 present a
case study for the use of the package in a scientific report. Section\ 5
concludes with a discussion and ideas for further development.

@@

hook(successes, k)
hook(quote(Ntotal), quote(N), quote=FALSE)
hook(prob, pi)
mathout(term)
```

To improve the situation, \pkg{mathml} provides a simple hook that can be used
to replace elements (e.g., verbose variable names) of the code by concise
mathematical symbols, as illustrated in the example. To simplify notation,
the `quote` flag of `hook()` defaults to TRUE, and `hook()` uses non-standard
evaluation to unpack its arguments. If quote is FALSE, as shown above, the user
has to provide the quoted expressions. Care should be taken to
avoid recursive hooks such as `hook(s, s["A"])` that endlessly replace
the \(s\) from \(s_{\mathrm{A}}\) as
in \(s_{\mathrm{A}_{\mathrm{A}_{\mathrm{A}\cdots}}}\).

Further customizations require the assertion of custom Prolog rules `math/2`,
`ml/3`, `jax/3`, as shown in the introduction.

@@

hook(ET(.Tau, .C, .MA, .SA, .MB, .SB, .MM), 
     E(`;`(T[.Tau], `,`(.MA, .SA, .MB, .SB, .MM))))
mathout(call("=", quote(E(T[tau])), ET))
```

@@

```{r, results="asis", echo=FALSE}
lsq <- function(tau, m, s, n)
{
  invisible(theta <- c(mu_A=0.53, sigma_A=4.3, mu_B=1.34, sigma_B=11.7, mu_M=160))
  optim(theta, gof, tau=tau, m=m, s=s, n=n, hessian=TRUE)
}

@@

```{r params, echo=FALSE}
fit <- lsq(tau, m, s, n)
theta <- fit$par
se <- sqrt(diag(solve(fit$hessian)))
cilo <- qnorm(0.025, theta, se)
ciup <- qnorm(0.975, theta, se)

@@

This package expands the possibilities of R\ by allowing calculations and
graphical rendering of the same terms. Building on current features of R\ and
existing packages for displaying formulas in
R\ [@murrell2000;@allaire2018]), the \pkg{mathml} package bridges the gap
between computational needs, presentation of results, and their reproducibility.
Two possible modes for typesetting are available to export R\ Markdown documents
in different formats or ShinyApp webpages: MathML and MathJax. A further, not-yet-mentioned, use of \pkg{mathml} is related to presentations made using
RMarkdown. By adopting \pkg{mathml} researchers or teachers can use the
RMarkdown to both conduct analysis and show results, smoothing the process and
uniforming the graphical output.

@@

In its current version \pkg{mathml} has some limitations. For example, it is
currently not possible to use the functions of \pkg{mathml} for writing
intratextual formulas; here, the user has to adopt the usual LateX notation.
However, we do not consider this to be a serious limitation, since the most
productive use of \pkg{mathml} still comes from more complicated calculations.

Also: In long equations, we did not yet find a convenient way to insert
linebreaks. This is mostly because it is not properly supported by MathML and
LaTeX renderers. In its current stage, the LaTeX package \pkg{breqn} [@breqn]
is mostly a proof of concept.

\pkg{mathml} is available for R\ version (…) and later, and can be easily
installed using the usual `install.packages("mathml")`. The source code of the
package is found at https://github.com/mgondan/mathml.

# Appendix

@@

# References
